---
title: "trnL_DB_Classification_and_Eval"
output: html_document
author: "Ozge Shera Kuddar"
---
```{r setup, include=FALSE}
library(reticulate)
knitr::opts_chunk$set(engine = "python", python.reticulate.autoprint = FALSE)
```

# Files:
#   Databases (fasta): ./DBs_fasta
#   Databases (taxonomy): ./DBs_taxonomy
#   Query datasets (fasta and taxonomy files) are located under ./Query


# Convert fasta files to DADA2 format for classification (created ./DBs_dada2 folder)

```{r, engine='python'} 
from Bio import SeqIO
import csv
import os
import sys

os.chdir("/Users/shera/trnL_DB_Classification_and_Eval")

taxonomy_dir = "./DBs_taxonomy"
fasta_dir = "./DBs_fasta"
output_dir = "./DBs_dada2"

os.makedirs(output_dir, exist_ok=True)

file_prefixes = [
    "Obitools_trnl_CD",
    "Obitools_trnl_CH",
    "Obitools_trnl_GH",
    "Rescript_trnl_CD",
    "Rescript_trnl_CH",
    "Rescript_trnl_GH",
    "Metacurator_trnl_CD",
    "Metacurator_trnl_CH",
    "Metacurator_trnl_GH"
]

for base_name in file_prefixes:

    current_csv_path = os.path.join(taxonomy_dir, f"{base_name}.csv")
    current_fasta_path = os.path.join(fasta_dir, f"{base_name}.fasta")
    current_output_path = os.path.join(output_dir, f"{base_name}_NBC.fasta")

    taxon_map_dict = {}

    with open(current_csv_path, 'r') as infile_csv:
        csv_reader_obj = csv.DictReader(infile_csv)
        for row in csv_reader_obj:
            acc_value = row["Acc"]
            taxon_items = [
                row["Kingdom"], row["Phylum"], row["Class"], row["Order"],
                row["Family"], row["Genus"], row["Species"]
            ]
            formatted_taxon = ";".join(taxon_items).replace(" ", "_")
            taxon_map_dict[acc_value] = formatted_taxon

    with open(current_output_path, 'w') as outfile_fasta:
        old_stdout = sys.stdout
        sys.stdout = open(os.devnull, "w")
        for seq_record in SeqIO.parse(current_fasta_path, "fasta"):
            original_id = seq_record.id
            if original_id in taxon_map_dict:
                seq_record.description = ""
                seq_record.id = taxon_map_dict[original_id]
                SeqIO.write(seq_record, outfile_fasta, "fasta")
        sys.stdout.close()
        sys.stdout = old_stdout

    print(f"Updated FASTA written: {current_output_path}")
```

# Use: Naive Bayes Classifier implemented in DADA2

```{R}
library(dada2)

set.seed(100) # reproducibility - which I did not use -shame on me- the files I generated, can be found @./Classified_BRC

query_dir <- "./Query"
ref_dir <- "./DBs_dada2"
out_dir <- "./Classified"
dir.create(out_dir, showWarnings = FALSE)

query_prefixes <- c(
  "Random_Combination",
  "Common_Species",
  "Mutated_Random_Combination",
  "Mutated_Common_Species"
)

regions <- c("CD", "CH", "GH")
tools <- c("Obitools", "Rescript", "Metacurator")

task_grid <- expand.grid(
  QueryPrefix = query_prefixes,
  Region = regions,
  Tool = tools,
  stringsAsFactors = FALSE
)

query_files <- file.path(
  query_dir,
  paste0(task_grid$QueryPrefix, "_", task_grid$Region, ".fasta")
)

ref_files <- file.path(
  ref_dir,
  paste0(task_grid$Tool, "_trnl_", task_grid$Region, "_NBC.fasta")
)

out_files <- file.path(
  out_dir,
  paste0(task_grid$QueryPrefix, "_", task_grid$Region, "_", task_grid$Tool, ".csv")
)

run_assignment <- function(query_fa, ref_fa, out_csv) {
  
  headers <- readLines(query_fa)
  mock_acc <- gsub("^>", "", headers[grepl("^>", headers)])
  
  taxa <- assignTaxonomy(query_fa, ref_fa, multithread = TRUE)
  df <- as.data.frame(taxa)
  
  df$Mock_Acc <- mock_acc
  df <- df[, c("Mock_Acc", setdiff(colnames(df), "Mock_Acc"))]
  
  write.csv(df, out_csv, row.names = FALSE)
  return(NULL)
}

mapply(
  FUN = run_assignment,
  query_fa = query_files,
  ref_fa = ref_files,
  out_csv = out_files,
  SIMPLIFY = FALSE
)
```


# Evaluation Step.1
# Assign Correctly classified (C), Misclassified (M), Unclassified (U) for Species, Genus, Family levels

```{r, engine='python'} 
import os
import pandas as pd

def evaluate_predictions():

    query_folder = "./Query" # Query taxonomy files (csv format)
    prediction_folder = "./Classified_BRC" # This folder contains the the files I generated (fasta format) you can change it to ./Classified
    output_folder = "./Eval1"
    os.makedirs(output_folder, exist_ok=True)

    taxonomy_levels = ["Species", "Genus", "Family"]

    # evaluation logic 
    def evaluate_row(row, mock_taxa):
        mock_acc = row["Mock_Acc"]

        if mock_acc not in mock_taxa:
            return {level + "_eval": "U" for level in taxonomy_levels}

        correct = mock_taxa[mock_acc]
        results = {}

        for level in taxonomy_levels:
            pred = row.get(level, "")
            corr = correct.get(level, "")

            if pred == "":
                results[level + "_eval"] = "U"
            elif pred == corr:
                results[level + "_eval"] = "C"
            else:
                results[level + "_eval"] = "M"

        return results

    # loop over prediction files
    for prediction_file in os.listdir(prediction_folder):
        if not prediction_file.endswith(".csv"):
            continue

        # remove _Tool_classified.csv
        core = prediction_file.rsplit("_", 2)[0]
        # e.g. Mutated_Common_Species_GH

        prefix, region = core.rsplit("_", 1)

        # Mutated queries use NON-mutated taxonomy
        if prefix.startswith("Mutated_"):
            taxonomy_prefix = prefix.replace("Mutated_", "", 1)
        else:
            taxonomy_prefix = prefix

        mock_csv = os.path.join(query_folder, f"{taxonomy_prefix}_{region}.csv")

        if not os.path.exists(mock_csv):
            print(f"[ERROR] Mock taxonomy file NOT found for: {prediction_file}")
            print(f"Expected: {mock_csv}")
            continue

        # load mock truth taxonomy
        mock_df = pd.read_csv(mock_csv, dtype=str).fillna("")
        mock_df["Acc"] = mock_df["Acc"].str.strip()
        mock_taxa = mock_df.set_index("Acc").to_dict(orient="index")

        # load predictions
        pred_path = os.path.join(prediction_folder, prediction_file)
        pred_df = pd.read_csv(pred_path, dtype=str).fillna("")

        pred_df["Mock_Acc"] = pred_df["Mock_Acc"].str.strip()

        # taxonomy formatting
        for c in taxonomy_levels:
            if c in pred_df.columns:
                pred_df[c] = pred_df[c].str.replace("_", " ").str.strip()

        # evaluate
        eval_df = pred_df.apply(lambda row: evaluate_row(row, mock_taxa), axis=1).apply(pd.Series)
        pred_df = pd.concat([pred_df, eval_df], axis=1)

        # save output
        out = prediction_file.replace(".csv", "_eval1.csv")
        pred_df.to_csv(os.path.join(output_folder, out), index=False)

        print(f"[OK] Evaluated: {prediction_file}  →  {out}")


if __name__ == "__main__":
    evaluate_predictions()
```


# Evaluation Step.2
# Generate Files for each query sets showing classification (C, M, U) and performance metrics (FC, P, R) 

```{r, engine='python'} 
import os
import pandas as pd
import sys

os.chdir("/Users/shera/trnL_DB_Classification_and_Eval")

eval_folder = "./Eval1"
output_folder = "./Eval2"
os.makedirs(output_folder, exist_ok=True)

tools = ["Obitools", "Rescript", "Metacurator"]

n_total = 3000  # total queries per dataset

# FC = (C + M) / n_total
# P  = C / (C + M)
# R  = C / n_total

summary_columns = [
    "Tool",
    # Species
    "Species_C", "Species_M", "Species_U",
    "Species_FC", "Species_P", "Species_R",
    # Genus
    "Genus_C", "Genus_M", "Genus_U",
    "Genus_FC", "Genus_P", "Genus_R",
    # Family
    "Family_C", "Family_M", "Family_U",
    "Family_FC", "Family_P", "Family_R"
]

all_files = [f for f in os.listdir(eval_folder) if f.endswith("_eval1.csv")]

groups = {}

for file in all_files:
    base = file.replace("_classified_eval1.csv", "")
    parts = base.split("_")

    tool = parts[-1]
    region = parts[-2]
    query = "_".join(parts[:-2])

    key = f"{query}_{region}"

    if key not in groups:
        groups[key] = {}
    groups[key][tool] = file

for key, tool_map in groups.items():

    rows = []

    for tool in tools:

        filename = tool_map.get(tool, None)

        if filename is None:
            sc = sm = su = 0
            gc = gm = gu = 0
            fc = fm = fu = 0
        else:
            df = pd.read_csv(os.path.join(eval_folder, filename))

            sc = (df["Species_eval"] == "C").sum()
            sm = (df["Species_eval"] == "M").sum()
            su = (df["Species_eval"] == "U").sum()

            gc = (df["Genus_eval"] == "C").sum()
            gm = (df["Genus_eval"] == "M").sum()
            gu = (df["Genus_eval"] == "U").sum()

            fc = (df["Family_eval"] == "C").sum()
            fm = (df["Family_eval"] == "M").sum()
            fu = (df["Family_eval"] == "U").sum()

        # Species metrics
        species_fc = round((sc + sm) / n_total, 3) if n_total > 0 else 0
        species_p  = round(sc / (sc + sm), 3) if (sc + sm) > 0 else 0
        species_r  = round(sc / n_total, 3) if n_total > 0 else 0

        # Genus metrics
        genus_fc = round((gc + gm) / n_total, 3) if n_total > 0 else 0
        genus_p  = round(gc / (gc + gm), 3) if (gc + gm) > 0 else 0
        genus_r  = round(gc / n_total, 3) if n_total > 0 else 0

        # Family metrics
        family_fc = round((fc + fm) / n_total, 3) if n_total > 0 else 0
        family_p  = round(fc / (fc + fm), 3) if (fc + fm) > 0 else 0
        family_r  = round(fc / n_total, 3) if n_total > 0 else 0

        row = {
            "Tool": tool,

            # Species
            "Species_C": sc, "Species_M": sm, "Species_U": su,
            "Species_FC": species_fc, "Species_P": species_p, "Species_R": species_r,

            # Genus
            "Genus_C": gc, "Genus_M": gm, "Genus_U": gu,
            "Genus_FC": genus_fc, "Genus_P": genus_p, "Genus_R": genus_r,

            # Family
            "Family_C": fc, "Family_M": fm, "Family_U": fu,
            "Family_FC": family_fc, "Family_P": family_p, "Family_R": family_r
        }

        rows.append(row)

    df_out = pd.DataFrame(rows, columns=summary_columns)

    output_path = os.path.join(output_folder, f"{key}.csv")
    df_out.to_csv(output_path, index=False)

    print(f"[OK] Created: {output_path}")
```


# Create summary file for each trnL region 

```{r, engine='python'} 
import os
import pandas as pd
import sys

os.chdir("/Users/shera/trnL_DB_Classification_and_Eval")

input_folder = "./Eval2"
output_folder = "./Eval2"
os.makedirs(output_folder, exist_ok=True)

regions = ["CD", "CH", "GH"]

for region in regions:

    combined_rows = []

    region_files = [f for f in os.listdir(input_folder) if f.endswith(f"_{region}.csv")]

    for filename in region_files:

        filepath = os.path.join(input_folder, filename)

        df = pd.read_csv(filepath)

        # extract Query_Set from filename
        # e.g., "Mutated_Common_Species_CD.csv" → "Mutated_Common_Species"
        query_set = filename.replace(f"_{region}.csv", "")

        df.insert(0, "Query_Set", query_set)

        combined_rows.append(df)

    # row-bind all 4×3 rows
    final_df = pd.concat(combined_rows, ignore_index=True)

    output_path = os.path.join(output_folder, f"trnL_{region}_eval.csv")
    final_df.to_csv(output_path, index=False)

    print(f"[OK] Created: {output_path}")
```

